First Order Alternation

### 2018-11-06 quantifiers/draft.tex
ok

@@

\begin{abstract}
\end{abstract}

\section{Preliminaries}

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
and false ($\bot$), respectively. The $\Data$ sort is equipped with
finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
where $\#(f)\geq0$ denotes the number of arguments (arity) of
$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
relation of arity $\#(p)$ and we write $\preds$ for the set of
predicates.

In the following, we shall consider that the interpretation of all
function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
predicates is fixed by the interpretation of the $\Data$ sort,
e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
are zero, the successor function and the arithmetic operations of
addition and multiplication. For simplicity, we further blur the
notational distinction between function symbols and their
interpretations.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. The size
$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
write it down.

A \emph{sentence} is a formula $\phi$ in which each variable occurs
under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
formula is \emph{positive} if each predicate symbol occurs under an
even number of negations and we denote by $\posforms(Q,X)$ the set of
positive formulae with predicates from the set $Q \subseteq \preds$
and free variables from the set $X \subseteq \vars$.

A formula is in \emph{prenex form} if it is of the form $\varphi =
Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
first-order formula can be written in prenex form, by renaming each
quantified variable to a unique name and moving the quantifiers
upfront.

%% For a variable $x \in \fv{}{\phi}$ and a term $t$, let $\phi[t/x]$ be
%% the result of replacing each occurrence of $x$ by $t$. For indexed
%% sets of terms $T=\set{t_1,\ldots,t_n}$ and variables
%% $X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
%% obtained by simultaneously replacing each occurrence of $x_i$ with
%% $t_i$ in $\phi$, for all $i\in[n]$.

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
case we say that $\phi$ \emph{entails} $\psi$.

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
of interpretations, a minimal element $\I \in \mathcal{S}$ is an
interpretation such that for no other interpretation $\I' \in
\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
and $\minsem{\phi}$ the set of minimal interpretations from
$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.

\section{First Order Alternating Automata}

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

Formally, a \emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
are the input data values that trigger the transition, whereas $q$ and
$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
in that state, respectively. Without loss of generality, we consider,
for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
most one such rule, as two or more rules can be joined using
disjunction.

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal. The original definition of alternating
finite-state automata \cite{ChandraKozenStockmeyer81} works around
this problem by considering boolean valuations (models) instead of
formulae.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$.

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
$T_i$ is a tree labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
  [n-1]$ in $T_i$, then the labels of its children form a cube from
  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
  a transition rule of $\A$.
\end{compactenum}
\end{definition}

\begin{definition}\label{def:accepting}
An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if \begin{compactitem}
\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
  and
  %
\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
  where $q \in F$.
\end{compactitem} 
If $\A$ has an accepting execution over $w$ starting with a cube $c
\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$.
\end{definition}
In this paper, we address the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

\subsection{Symbolic Execution}

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence. Otherwise stated,
$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
a word $w$ such that $\event{w} = \alpha$.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. As a
shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
when no confusion arises.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ is defined as:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
,~\forall q\in Q ~\forall i \in [n]\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
  subformula of $\pathform{\event{w}}$ corresponding to the
  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
  interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
  \data{w} \models \pathform{\event{w}}$, by induction on $n \geq
  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
  \data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
  \rightarrow \stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the
  execution semantics of $\A$. This is the case because the children
  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
  $(n-1)$-th level of $\T$ form a cube from
  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
  each $y_i$ the value $d_i$ and behaves like $\data{w}$,
  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
  that $\J,\data{w} \models \pathform{\event{w}}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs on the
  right-hand side. Then it must be the case that, for some node on the
  $(n-1)$-th level of $\G$, labeled with a configuration
  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}

Next, we give a logical characterization of acceptance, relative to a
given sequence of input events $\alpha \in \Sigma^*$. To this end, we
constrain the path formula $\pathform{\alpha}$ by requiring that only
final states of $\A$ occur on the last level of the execution.  The
result is the \emph{acceptance formula} for $\alpha$:
\begin{equation}\label{eq:accform}
\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{n}(\vec{y}) \rightarrow \bot
\end{equation}
The top-level universal quantifiers from a subformula $\forall y_1
\ldots \forall y_{\#(q)} ~.~ \stamp{q}{i}(\vec{y}) \rightarrow \psi$
of $\accform{\alpha}$ will be referred to as \emph{path quantifiers},
in the following. Notice that path quantifiers are distinct from the
transition quantifiers that occur within a formula $\psi$ of a
transition rule $q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi$ of $\A$.

The acceptance formula $\accform{\A}$ is false in every interpretation
of the predicates that assigns a non-empty set to a non-final
predicate occurring on the last level in the execution forest. The
relation between the words accepted by $\A$ and the acceptance formula
above, is formally captured by the following lemma:

\begin{lemma}\label{lemma:acceptance}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
  every word $w \in \Sigma[X]^*$, the following are
  equivalent: \begin{compactenum}
  \item\label{it1:lemma:acceptance} there exists an interpretation
    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$, 
    %
  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
  \end{compactenum}
\end{lemma}
\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
  that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
  that \begin{inparaenum}[(i)]
    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
      that
    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
      final configurations only. \end{inparaenum} First, assume that
  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
  m < n$. Then there exists a node on the $m$-th level, labeled with
  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
  children. By the definition of the execution semantics of $\A$, we
  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
  \data{w}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
  $\I,\data{w} \models \accform{\alpha}$, we obtain that $\I,\eta \models
  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
  contradicts the fact that $\I=\I_{\T}$ and that
  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
  \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
  similar reasoning as in the above case, we obtain that
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.

  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
  of $\A$ over $w$. We prove that $\I_\T,\data{w} \models
  \accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
  $\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
  labeled by final configurations, we obtain that $\I_\T,\data{w}
  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
  trivially.  \qed}

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely \begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

To tackle this problem, we start from the observation that acceptance
formulae have a particular form, which allows the elimination of path
quantifiers and of predicates, by a couple of
satisfiability-preserving transformations. The result of applying
these transformations is a formula with no predicate symbols, whose
only quantifiers are those introduced by the transition rules of the
automaton, referred to as transition quantifiers. We shall further
assume (\S\ref{sec:emptiness}) that the first-order theory of the data
sort $\Data$ has quantifier elimination, which allows to effectively
decide the satisfiability of such formulae.

For the time being, let us formally define the elimination of
transition quantifiers and predicates, respectively. Consider a given
sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
where $\alpha_0=\epsilon$.

\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
Observe that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

\begin{example}\label{ex:quant-elim}
  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
  \[\begin{array}{rcl}
  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
  \Delta & = & \set{
    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
  }
  \end{array}\]
  For the input event sequence $\alpha = a_1a_2$, the acceptance
  formula is:
  \[\begin{array}{rcl}
  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
  \end{array}\]
  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
  \[\begin{array}{rcl}
  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)] 
  \enspace\enspace\hfill\blacksquare
  \end{array}\]
\end{example}

The next lemma establishes a formal relation between the
satisfiability of an acceptance formula $\accform{\alpha}$ and that of
the formula $\quantform{\alpha}$, obtained by eliminating the path
quantifiers from $\accform{\alpha}$.

\begin{lemma}\label{lemma:quant}
  For any input event sequence $\alpha=a_1\ldots a_n$ and each
  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
  hold: \begin{compactenum}
  \item\label{it1:quant} for all interpretations $\I$, if $\I,\nu
    \models \accform{\alpha}$ then $\I,\nu \models
    \quantform{\alpha}$. 
    %
  \item\label{it2:quant} if there exists an interpretation $\I$ such
    that $\I,\nu \models \quantform{\alpha}$ then there exists an
    interpretation $\J \subseteq \I$ such that $\J,\nu \models
    \accform{\alpha}$. 
  \end{compactenum}
\end{lemma}
\proof{(\ref{it1:quant}) Trivial, since every subformula
  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.

  \noindent(\ref{it2:quant}) By repeated applications of the following fact:

>>>>

\renewcommand{\baselinestretch}{0.99}

@@

\input abstract
\input body
